---
title: STATS531 Midterm Project \\n Time Series Analysis on Fatal Car Accidents in
  Michigan
author: "Geovanna Caballero, Annie Tai, Paul Zhang"
date: "2025-02-11"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction





### Data Description
The data for this study is obtained from the National Highway Traffic Safety Administration (NHTSA) and covers the years 2008 to 2022 [1]. The data is aggregated on a monthly basis and includes the number of fatal crashes and the number of people killed in fatal crashes in both Michigan and the overall United States. 



### Exploratory Analysis



```{r preprocessing, include=FALSE, echo=FALSE}
library(readxl)
library(tidyverse)
library(lubridate)
library(tseries)
library(forecast)

process_data <- function(pathname, skip){
  df <- read_excel(pathname, skip=skip)
  #removing total row and total column
  df <- df %>%
    slice_head(n = -1) %>% 
    select(-last_col())
  
  colnames(df)[1] <- "Year"
  #pivot table so each row is crashes in a given month & year
  df_long <- df %>%
    pivot_longer(cols = -Year, names_to = "Month", values_to = "Crashes")

  #creating date column
  df_long <- df_long %>% 
    mutate(Month = match(Month, month.name)) %>% 
    mutate(Date = make_date(Year, Month, 1)) %>%  
    arrange(Date) 
  return(df_long)
}
crashes_MI <- process_data("CrashReport_FatalCrashes_MI.xlsx",  skip=7)
deaths_MI <- process_data("CrashReport_KilledInFatalCrashes_MI.xlsx",  skip=8)
# crashes_US <- process_data("CrashReport_FatalCrashes_US.xlsx", skip=6)
# deaths_US <- process_data("CrashReport_KilledInFatalCrashes_US.xlsx", skip=7)
# crashes_MI_filtered <- process_data("./midterm_project/CrashReport_FatalCrashes_MI_Interstate_Speeding_Darkcondition.xlsx", skip=8)

# Convert the tables to series. The code is borrowed from the Lecture Notes.
table2ts = function(data, start = c(2008, 1)){
    data_ts = ts(data, start = start, frequency = 12)
    return(data_ts[, "Crashes"])
}

crashes_MI_ts = table2ts(crashes_MI)
deaths_MI_ts = table2ts(deaths_MI)
# crashes_US_ts = table2ts(crashes_US)
# deaths_US_ts = table2ts(deaths_US)

date = seq(from = 2008, length = length(crashes_MI_ts), by = 1/12)

```

#### Some Descriptive Statistics
```{r}
psych::describe(crashes_MI_ts)
psych::describe(deaths_MI_ts)
# psych::describe(crashes_US_ts)
# psych::describe(deaths_US_ts)
```

```{r, fig.align = "center", fig.dim = c(12, 12)}
par(mfrow = c(2, 1))
plot(date, crashes_MI_ts, type = "l")
plot(date, deaths_MI_ts, type = "l")
# plot(date, crashes_US_ts, type = "l")
# plot(date, deaths_US_ts, type = "l")
```


### Model Selection
#### Stationarity Tests

```{r}
adf.test(crashes_MI_ts)
adf.test(deaths_MI_ts)
```
Both time series have passed the ADF test. 

__Updates__ 
I have found another statistical test called the KPSS Test^[https://en.wikipedia.org/wiki/KPSS_test]. It seems to be an advanced version of 
stationary test since it takes trends into consideration. The trend doesn't have to be linear.

The KPSS test may be helpful since our data apparently have nonlinear trends.

The null hypothesis, the alternative hypothesis for the test are as follows:

  + $H_0$: The time series is a trend-stationary process (A stochastic process from which an underlying trend 
(function solely of time) can be removed, leaving a stationary process^[https://en.wikipedia.org/wiki/Trend-stationary_process].)
  + $H_1$: The time series is a unit root process.

```{r}
kpss.test(crashes_MI_ts, null = "Trend")
kpss.test(deaths_MI_ts, null = "Trend")
```
Both of $p$-values exceed 0.05, indicating trend stationary.

#### ARMA Models, Before Detrending
The following code block generates the AIC table, given a stationary time series. It's borrowed from the lecture notes^[https://ionides.github.io/531w25/05/slides.pdf, pp.29].
```{r}
aic_table <- function(data,P,Q){
table <- matrix(NA,(P+1),(Q+1))
for(p in 0:P) {
    for(q in 0:Q) {
    # table[p+1,q+1] <- arima2::arima(data,order=c(p,0,q))$aic
    table[p + 1, q + 1] = tryCatch({
        arima(data, order = c(p, 0, q))$aic},
        error = function(e) {NA}
        )
    }
}
dimnames(table) <- list(paste("AR",0:P, sep=""),
paste("MA",0:Q,sep=""))
table
}
require(knitr)
```

##### Crashes
```{r, warning=FALSE}
crashes_table = aic_table(crashes_MI_ts, 4, 3)
kable(crashes_table, digits=2)
```

ARMA(2, 3) vs ARMA(4, 3). The AIC's are close.

```{r}
crashes_arma23 = arima(crashes_MI_ts, order = c(2, 0, 3))
crashes_arma43 = arima(crashes_MI_ts, order = c(4, 0, 3))
```

__Likelihood Ratio Test__
We have two nested parameter spaces and therefore we can try LRT.
```{r}
lrt = function(model0, model1, df){
    chi_sq = 2 * (model1$loglik - model0$loglik)
    pval = pchisq(chi_sq, df, lower.tail = FALSE)
    cat(sprintf("Test Statistic: %.4f\nDOF: %d\np-value: %.4f", chi_sq, df, pval))
    }
```

```{r}
lrt(crashes_arma23, crashes_arma43, 2)
```

Fitting ARMA(4, 3) is not necessary.

 __Diagnostics for ARMA(2,3)__:

+ Check the AR roots (the code also comes from the lecture notes^[https://ionides.github.io/531w25/05/slides-annotated.pdf, p.32]):
```{r}
AR_roots <- polyroot(c(1,-coef(crashes_arma23)[c("ar1","ar2")]))
abs(AR_roots)
```
That's an edge case. 

+ Ljung-Box Test
```{r}
Box.test(crashes_arma23$residuals, lag = 20, type = "Ljung-Box")
```

+ Residual Plots
```{r, fig.align = "center", fig.dim = c(12, 3)}
par(mfrow = c(1, 3))
plot(crashes_arma23$residuals)
acf(crashes_arma23$residuals)
qqnorm(crashes_arma23$residuals, pch = 1, frame = FALSE)
qqline(crashes_arma23$residuals, col = "steelblue", lwd = 2)
```

positive autocorrelation; right skew


##### Deaths
The deaths series.
```{r,warning=FALSE}
deaths_table = aic_table(deaths_MI_ts, 4, 3)
kable(deaths_table, digits=2)
```
ARMA(3, 2)

```{r}
deaths_arma32 = arima(deaths_MI_ts, order = c(3, 0, 2))
summary(deaths_arma32)
```

__Diagnostics__

+ Check AR roots
```{r}
AR_roots <- polyroot(c(1,-coef(deaths_arma32)[c("ar1","ar2","ar3")]))
abs(AR_roots)
```

+ Ljung-Box Test
```{r}
Box.test(deaths_arma32$residuals, lag = 20, type = "Ljung-Box")
```

+ Residual Plots
```{r, fig.align = "center", fig.dim = c(12, 3)}
par(mfrow = c(1, 3))
plot(deaths_arma32$residuals)
acf(deaths_arma32$residuals)
qqnorm(deaths_arma32$residuals, pch = 1, frame = FALSE)
qqline(deaths_arma32$residuals, col = "steelblue", lwd = 2)
```

light-tailed on the left

#### ARMA models, After Detrending
Let's see what happens after we detrend the data. (I read some docs^[https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/stl] be4 writing the following code.)
```{r}
crashes_MI_decomposed = stl(crashes_MI_ts, s.window = "periodic")
crashes_MI_detrended = crashes_MI_ts - crashes_MI_decomposed$time.series[, "trend"]

deaths_MI_decomposed = stl(deaths_MI_ts, s.window = "periodic")
deaths_MI_detrended = deaths_MI_ts - deaths_MI_decomposed$time.series[, "trend"]

```

##### Crashes, detrended
ARMA(2, 2) has the lowest AIC, but ARMA(2, 3) is close.
```{r, warning=FALSE}
crashes_detrend_table = aic_table(crashes_MI_detrended, 4, 3)
kable(crashes_detrend_table, digits=2)
```
```{r}
crashes_detrend_arma22 = arima(crashes_MI_detrended, order = c(2, 0, 2))
crashes_detrend_arma23 = arima(crashes_MI_detrended, order = c(2, 0, 3))
```


__Likelihood Ratio Tests__
```{r}
lrt(crashes_detrend_arma22, crashes_detrend_arma23, 1)
```

ARMA(2, 2) is good.

__Diagnostics__

+ Check the AR roots
```{r}
AR_roots <- polyroot(c(1,-coef(crashes_detrend_arma22)[c("ar1","ar2")]))
abs(AR_roots)
```

+ Ljung-Box Test
```{r}
Box.test(crashes_detrend_arma22$residuals, lag = 20, type = "Ljung-Box")
```

+ Residual Plots
```{r, fig.align = "center", fig.dim = c(12, 3)}
par(mfrow = c(1, 3))
plot(crashes_detrend_arma22$residuals)
acf(crashes_detrend_arma22$residuals)
qqnorm(crashes_detrend_arma22$residuals, pch = 1, frame = FALSE)
qqline(crashes_detrend_arma22$residuals, col = "steelblue", lwd = 2)
```


##### Deaths, detrended
ARMA(2, 3)
```{r, warning=FALSE}
deaths_detrend_table = aic_table(deaths_MI_detrended, 4, 3)
kable(deaths_detrend_table, digits=2)
```

```{r}
deaths_detrend_arma23 = arima(deaths_MI_detrended, order = c(2, 0, 3))
```

__Diagnostics__

+ Check the AR roots
```{r}
AR_roots <- polyroot(c(1,-coef(deaths_detrend_arma23)[c("ar1","ar2")]))
abs(AR_roots)
```

+ Ljung-Box Test
```{r}
Box.test(deaths_detrend_arma23$residuals, lag = 20, type = "Ljung-Box")
```

+ Residual Plots
```{r, fig.align = "center", fig.dim = c(12, 3)}
par(mfrow = c(1, 3))
plot(deaths_detrend_arma23$residuals)
acf(deaths_detrend_arma23$residuals)
qqnorm(deaths_detrend_arma23$residuals, pch = 1, frame = FALSE)
qqline(deaths_detrend_arma23$residuals, col = "steelblue", lwd = 2)
```




### Conclusion



### References
[1] https://cdan.dot.gov/query
[2] https://ionides.github.io/531w25/08/slides.pdf, p.6
