---
title: "**STATS 531 Homework 7**"
author: "Paul Zhang"
output: pdf_document
# fontsize: 6pt
---
```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE,  message = FALSE) # supressing warnings 
```

# 0. Libraries and Data
```{r}
library(doParallel)
```

# 1. Greatlakes Cluster Test
## Feedbacks

+ The test code runs almost perfectly on the server. The only issue I have encountered so far is that the waiting time before my batch job starts can be long sometimes
+ For the record, I am familiar with Linux, but I have not used any Linux clusters before.
+ For some reason, the code sometimes runs significantly faster on my laptop, which seems suspicious. To investigate whether the issue is related to parallelization, 
I modified `test.R` to include the number of cores used. The modified file can be found in the Appendix.

Below are the runtimes for the test code on the server and my laptop. 

__Greatlakes Cluster__
```{bash, eval = FALSE}
[paulzh@gl-login3 hw7]$ cat test.csv
"","user.self","sys.self","elapsed","user.child","sys.child","cores"
"time0",5.224,0.29,5.527,0,0,36
"time1",0.124,0.826,1.777,0.53,0.122,36
"time2",0.102,0.748,1.172,1.816,0.511,36
"time3",0.239,0.578,1.166,5.422,1.75,36
"time4",0.969,0.877,2.044,5.539,2.277,36
```

__Local Env__
```{bash, eval=FALSE}
[paulzhang@ubuntu Homework7(main)]$ cat test.csv
"","user.self","sys.self","elapsed","user.child","sys.child","cores"
"time0",1.753,0.189,1.942,0,0,32
"time1",0.108,0.501,1.311,0.273,0.1,32
"time2",0.089,0.571,1.178,5.383,1.495,32
"time3",0.089,0.662,1.286,5.538,2.067,32
"time4",0.617,1.215,2.419,5.433,6.726,32
```




## Stop Typing SSH Addresses Over and Over
Manually entering SSH login addresses each time is infuriating. Fortunately, this can be easily fixed by adding the following entry to 
`~/.ssh/config`:
```{bash, eval = FALSE}
Host gl
    HostName greatlakes.arc-ts.umich.edu
    User uniqname

Host glxfer
    HostName greatlakes-xfer.arc-ts.umich.edu
    User uniqname
```
After that, the following commands
```{bash, eval = FALSE}
ssh uniqname@greatlakes.arc-ts.umich.edu
scp myfile uniqname@greatlakes-xfer.arc-ts.umich.edu:mydir
```
can be replaced by 
```{bash, eval = FALSE}
ssh gl
scp myfile glxfer:mydir
```


# 2. Likelihood Maximization for SEIR

